Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-04-10 brms_meeg.qmd
updating the manuscript (intro)

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
abstract: "Time-resolved electrophysiological measurements such as those obtained through magneto- or electro-encephalography (M/EEG) offer a unique window into the neural activity underlying cognitive processes. Researchers are often interested in determining whether and when these signals differ across experimental conditions or participant groups. The conventional approach involves mass-univariate statistical testing across time, followed by corrections for multiple comparisons (e.g., FDR, FWER) or cluster-based inference. While effective for controlling error rates, cluster-based methods come with a significant limitation: they shift the focus of inference from individual time points to clusters, making it difficult to draw precise conclusions about the onset or offset of observed effects. Here, we introduce a *model-based* approch for analysing M/EEG timeseries such as event-related potentials (ERPs) or decoding performance over time. Our method leverages Bayesian generalised additive multilevel models (GAMMs), providing posterior probabilities that an effect is above zero (or above chance) at each time point, while naturally accounting for temporal dependencies and between-subject variability. Using both simulated and actual M/EEG datasets, we demonstrate that this approach substantially outperforms conventional methods in estimating the onset and offset of neural effects, yielding more precise and reliable results. We also provide an R package implementing the method and show how it can be seamlessly integrated into M/EEG analysis pipelines using MNE-Python."
keywords: [EEG, MEG, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(tidytext)
library(grateful)
library(ggrepel)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)

# setting knitr options
knitr::opts_chunk$set(
    cache = TRUE,
    eval = TRUE,
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    out.width = "100%",
    fig.asp = 0.75,
    fig.pos = "!htb",
    # dev = "cairo_pdf",
    dev = "png",
    dpi = 300
    )

# setting a default ggplot2 theme
# theme_set(theme_light(base_size = 10, base_family = "Open Sans") )
theme_set(theme_bw(base_size = 10, base_family = "Open Sans") )

# template for citing R packages
# pkrt_set(pkg = "the `R` package `:pkg` [:ref]")
pkrt_set(pkg = "the `R` package `:pkg` v :ver [:ref]")
```

@@

Understanding the temporal dynamics of cognitive processes requires methods that can capture fast-changing neural activity with high temporal resolution. Magnetoencephalography (MEG) and electroencephalography (EEG) are two such methods, widely used in cognitive neuroscience for their ability to track brain activity at the millisecond scale. These techniques provide rich time series data that reflect how neural responses unfold in response to stimuli or tasks. A central goal in many M/EEG studies is to determine whether and when neural responses differ across experimental conditions or participant groups.

Here are some useful references to be discussed [@combrisson_exceeding_2015; @ehinger_unfold_2019; @hayasaka_validating_2003; @luck_how_2017; @frossard2021; @frossard2022; @gramfort2013; @maris2007; @pernet2015]... See also @maris2011 and @rosenblatt2018 (history of cluster-based approaches)... Clusters failures [@eklund2016]...

@@

Description of cluster-based approaches [see @sassenhagen2019]...

However, as pointed by the original authors themselves [@maris2007], "there is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA \[false alarm\] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect" [@maris2007; @sassenhagen2019]. In contrast, the proposed approach, based on Bayesian generalised additive multilevel models, allows quantifying the posterior probability of effects being above chance at the level of timesteps, voxels, sensors (etc), while naturally taking into account spatiotemporal dependencies present in M/EEG timeseries.

## Related work

@@

@@

Recently, @teichmann2022 provided a detailed tutorial on using Bayes factors (BFs) to analyse the 1D or 2D output from MVPA, that is, for testing, at every timestep, whether decoding performance is above chance level. However, this approach provides timeseries of BFs that ignores temporal dependencies..

## Bayesian regression modelling

Short intro/recap about Bayesian (linear and generalised) regression models...

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. To clarify the terminology at this stage: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., a cubic polynomial) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions. Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function. In other words, more complex (wiggly) basis function are penalised.

@@

Introduction to multilevel GAMs [@pedersen_hierarchical_2019]...

Now describe the Bayesian GAMM [@miller2025]... Proper inclusion of varying/random effects in the model specification protects against overly wiggly curves [@baayen2020]... Generalising to scale and shape or "distributional GAMs" [@rigby2005; @umlauf2018] and applied to neuroimaging data [@dinga2021]...

@@

Given the previously reported limitations of conventional methods to precisely identify the onset and offset of M/EEG effects (e.g., ERPs, decoding performance), we aimed to provide a model-based approach for estimating the onset and offset of such effects. To achieve this, we leveraged Bayesian generalised additive multilevel models (BGAMMs) fitted via the `brms R` package and compared the performance of this approach to conventional methods on both simulated and actual M/EEG data.

@@

Following the approach of @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see code in the online supplementary materials and details in @yeung2004]. As in @rousselet_using_2025, the signal was generated from truncated Gaussian with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated data for a group of 20 participants with 50 trials per participant and condition (@fig-eeg).

@@

# importing helper functions
# source("code/functions.R")

@@

# plotting the data
raw_df %>%
    summarise(
        eeg_mean = mean(eeg),
        eeg_se = sd(eeg) / sqrt(n() ),
        .by = c(participant, condition, time),
        ) %>%
    ggplot(aes(x = time, y = eeg_mean, colour = condition, fill = condition) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_ribbon(
        aes(ymin = eeg_mean - eeg_se, ymax = eeg_mean + eeg_se, colour = NULL),
        alpha = 0.3, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~participant) +
    scale_colour_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")
```

````{=html}
<!--

We computed the average of the ERP difference (@fig-erp)...

```{r fig-erp, eval = FALSE, echo = FALSE, out.width = "75%", fig.cap = "Group-level average difference between conditions (mean +/- standard error of the mean). The 'true' onset and offset are indicated by the vertical dashed lines."}
# averaging across participants
raw_df %>%
    pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
    mutate(eeg_diff = cond2 - cond1) %>%
    summarise(
        eeg_mean = mean(eeg_diff),
        eeg_sem = sd(eeg_diff) / sqrt(n() ),
        .by = time
        ) %>%
    mutate(lower = eeg_mean - eeg_sem, upper = eeg_mean + eeg_sem) %>%
    # plotting the data
    ggplot(aes(x = time, y = eeg_mean) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_ribbon(
        aes(ymin = lower, ymax = upper, colour = NULL),
        alpha = 0.3, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    labs(x = "Time (s)", y = "EEG difference (a.u.)")
```

-->
````

## Model fitting

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms` (i.e., weakly informative priors). We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$ [@gabry2019; @gelman2020]. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects. @fig-plot-post-slope shows the predictions of this model together with the raw data.

@@

# timing the model fitting
# around 600s for 2000 iterations on 4 chains and 4 cores
# around 190s for 2000 iterations on 4 chains and 4 cores + cmdstanr and 01
# around 440s for 2000 iterations on 4 chains and 4 cores + threading(2), cmdstanr and 01
# library(tictoc)
# tic()

# fitting the GAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    # eeg ~ condition + s(time, bs = "cr", k = 20, by = condition),
    # thin-plate regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "tp", k = 20, by = condition),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/gam.rds"
    #################################
    # speeding up the model fitting #
    #################################
    # backend = "cmdstanr",
    # threads = threading(2),
    # stan_model_args = list(stanc_options = list("O1") )
    )

# stop timing
# toc()
```

@@

```{r model-preds, echo = FALSE}
# defining a function to plot posterior predictions
plot_post_preds <- function (model, data = NULL) {

@@

    post_preds %>%
        ggplot(aes(x = time, y = eeg, colour = condition, fill = condition) ) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_vline(xintercept = true_offset, linetype = 2) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_point(
            data = data %>% summarise(eeg = mean(eeg), .by = c(time, condition) ),
            aes(fill = NULL),
            pch = 21, show.legend = FALSE
            ) +
        geom_line(aes(y = estimate__), show.legend = FALSE) +
        geom_ribbon(
            aes(ymin = lower__, ymax = upper__, colour = NULL),
            alpha = 0.25, show.legend = FALSE
            ) +
        scale_colour_met_d(name = "Johnson") +
        scale_fill_met_d(name = "Johnson") +
        labs(x = "Time (s)", y = "EEG signal (a.u.)")

}
```

```{r gam-preds, echo = FALSE, eval = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior predictions of the GAM (left) and GP (right) models."}
# plotting the posterior predictions
plot_post_preds(model = gam) + plot_post_preds(model = gp_model)
```

We depict the posterior predictions together with the posterior estimate of the slope for `condition` at each timestep (@fig-plot-post-slope).

>>>>

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
abstract: "Time-resolved electrophysiological measurements such as those obtained through magneto- or electro-encephalography (M/EEG) offer a unique window into the neural activity underlying cognitive processes. Researchers are often interested in determining whether and when these signals differ across experimental conditions or participant groups. The conventional approach involves mass-univariate statistical testing across time and space, followed by corrections for multiple comparisons such as cluster-based inference. While effective for controlling error rates at the cluster-level, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual time points to clusters, it makes difficult to draw precise conclusions about the onset or offset of observed effects. Here, we introduce a *model-based* approach for analysing M/EEG timeseries such as event-related potentials (ERPs) or decoding performance over time. Our method leverages Bayesian generalised additive multilevel models (GAMMs), providing posterior probabilities that an effect is above zero (or above chance) at each time point, while naturally accounting for temporal dependencies and between-subject variability. Using both simulated and actual M/EEG datasets, we demonstrate that this approach substantially outperforms conventional methods in estimating the onset and offset of neural effects, yielding more precise and reliable results. We provide an R package implementing the method and show how it can be seamlessly integrated into M/EEG analysis pipelines using MNE-Python."
keywords: [EEG, MEG, cluster-based inference, simulation, multiple comparisons, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
          \usepackage{booktabs, caption, longtable, colortbl, array}
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(tidytext)
library(grateful)
library(ggrepel)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)
library(mgcv)
library(gt)

# setting knitr options
knitr::opts_chunk$set(
    cache = TRUE,
    eval = TRUE,
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    out.width = "100%",
    fig.asp = 0.75,
    fig.pos = "!htb",
    dev = "png",
    dpi = 300
    )

# setting a default ggplot2 theme
theme_set(theme_bw(base_size = 10, base_family = "Open Sans") )

# template for citing R packages
pkrt_set(pkg = "the `R` package `:pkg` v :ver [:ref]")
```

@@

@@

Understanding the temporal dynamics of cognitive processes requires methods that can capture fast-changing neural activity with high temporal resolution. Magnetoencephalography and electroencephalography (M/EEG) are two such methods, widely used in cognitive neuroscience for their ability to track brain activity at the millisecond scale. These techniques provide rich time series data that reflect how neural responses unfold in response to stimuli or tasks. A central goal in many M/EEG studies is to determine whether, when, and where neural responses differ across experimental conditions or participant groups.

The conventional approach involves mass-univariate statistical testing with a unique threshold, followed by corrections for multiple comparisons with the goal of maintaining the familywise error rate (FWER) or false discovery rate (FDR) at the nominal level. Cluster-based inference is the most common way of achieving this sort of error control in the M/EEG literature, being the default option in several software programs. While effective for controlling error rates, cluster-based inference comes with a significant limitation: they shift the focus of inference from individual time points to clusters, making it difficult to draw precise conclusions about the onset or offset of observed effects [@maris2007; @sassenhagen2019]. As pointed by @maris2007; "there is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA \[false alarm\] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect". Even worse, as noted by @rosenblatt2018, cluster-based inference suffers from low spatial resolution: "Since discovering a cluster means that 'there exists at least one voxel with an evoked response in the cluster', and not that 'all the voxels in the cluster have an evoked response', it follows that the larger the detected cluster, the less information we have on the location of the activation." As a consequence, cluster-based inferenc has been shown...

To overcome the limitations of cluster-based inference, we introduce a novel model-based approach for precisely localising M/EEG effects in time (space, and other dimensions). The proposed approach, based on Bayesian generalised additive multilevel models, allows quantifying the posterior probability of effects being above chance at the level of timesteps, sensors, voxels, etc., while naturally taking into account spatiotemporal dependencies present in M/EEG timeseries. We compare the performance of the proposed approach to well-established alternative methods using both simulated and actual M/EEG data and show that it significantly outperforms alternative methods in estimating the onset and offset of M/EEG effects.

## Cluster-based inference

Different methods exist to control the family-wiser error rate (FWER), defined as the type-1 error rate (false positive) over an ensemble (family) of tests... for instance the Bonferroni correction [@dunn1961], however this method is generally overconservative as it assumes statistical independence of tests, an assumptions that is clearly violated in the context of M/EEG timeseries with spatiotemporal dependencies... or the false discovery rate (FDR), defined as the proportion of false positive among positive tests [e.g., @benjamini1995; @benjamini2001]...

<!--

Here are some useful references to be discussed [@ehinger_unfold_2019; @luck_how_2017; @frossard2021; @frossard2022]...

-->

On popular technique to account for spatiotemporal dependencies while controlling the FWER is cluster-based inference. Description of cluster-based approaches [see @sassenhagen2019]... A typical cluster-based inference consists of two successive steps. First, clusters are defined as sets of contiguous voxels, channels, and/or timesteps, whose intensity/activity exceeds some predefined threshold. Clusters are then characterised by their height (i.e., maximal value), extent (number of constituent elements), or some combination of both, such at its "mass", for instance by summing the statistics within a cluster, an approach refereed to as "cluster mass" [@maris2007; @pernet2015]. Then, the null hypothesis is tested by assessing the probability.. As different cluster-forming thresholds lead to clusters with different spatial or temporal extent, this initial threshold modulates the sensitivity of the subsequent permutation test. The threshold-free cluster enhancement method (TFCE) was introduced by @smith2009 to overcome this choice of an arbitrary threshold.

where $h_{0}$ is typically $0$ and parameters $E$ and $H$ are set a priori (typically to 0.5 and 2, respectively) and control the influence of the extend and height on the TFCE. Then, p-value for timestep/voxel $t$ is computed by comparing it TFCE with the null distribution of TFCE values. For each permuted signal, we keep the maximal value over the whole signal for the null distribution of the TFCE. The TFCE combined with permutation (assuming a large enough number of permutations) has been shown to provide accurate type 1 FWER [e.g., @pernet2015]. However, previous simulation work showed that cluster-based methods have indeed poor performance in localising the onset of M/EEG effects [e.g., @rousselet_using_2025; @sassenhagen2019].

To sum up this section, cluster-based inference main limitations is that it provides inference at the cluster level only, not allowing inference at the level of timesteps, sensors, etc. Thus, it does not allow inferring the localisation of effects [for more details on cluster-based inference, see for instance @maris2007; @maris2011; @sassenhagen2019]. In the following, we briefly review previous modelling work. Then, we provide a brief introduction to Bayesian statistical modelling, before moving to generalised additive models (GAMs) and generalised additive multilevel models (GAMMs) to illustrate how they can be used to precisely localise the onset and offset of M/EEG effects.

## Previous work on modelling M/EEG data

@@

As put by @rousselet_using_2025, concluding on the onset of effect based on a series of univariate tests... commits to three fallacies... here, we want to avoid these by introducing a model-based approach, which naturally take into account the temporal dependencies in the data to output a series of posterior probabilities...

@@

Recently, @teichmann2022 provided a detailed tutorial on using Bayes factors (BFs) to analyse the 1D or 2D output from MVPA, that is, for testing, at every timestep, whether decoding performance is above chance level. However, this approach provides timeseries of BFs that ignores temporal dependencies...

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. To clarify the terminology at this stage: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., cubic polynomial, thin-plate) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions (@fig-intro-gam). Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function. In other words, more complex (wiggly) basis function are penalised.

@@

The Bayesian approach to statistical modelling is characterised by its reliance on probability theory to [@gelman2020]... In this framework, all unknown entities are assigned probability distributions reflecting the uncertainty... These probability distributions are commonly referred to as "priors" and represent some state of knowledge about unknown quantities before seeing any data. There are debates among Bayesian practitioners as to whether prior distributions should encore subjective (personal) beliefs or... but these debates are outside the scope of the present paper and we therefore the interested reader to dedicated work (e.g., XX; YY)... In practice, weakly informative priors are often used as default priors in situations in which subjective priors are difficult to define/elicit... Bayesian models are then fitted on empirical (actual or simulated) data to update prior states of knowledge to posterior states of knowledge using Bayes theorem, or in practise, sampling-based approximations of the posterior distribution...

```{r fig-intro-gam, echo = FALSE, fig.width = 9, fig.cap = "Different types of GAM(M)s. **A**: GAMs predictions are computed as the weigthed sum (in black) of basis functions (here thin-plate basis functions, in grey). **B**: Constant-effect GAM, with 5 participants in colours and the group-level prediction in black. **C**: Varying-intercept + varying-slope GAMM (with common smoother). **D**: Varying-intercept + varying-slope + varying-smoother GAMM. In this model, each participant gets its own intercept, slope, and degree of 'wiggliness' (smoother)."}
########################
# simulating som edata #
########################

set.seed(666)
n_subjects <- 5
n_per_subject <- 50
subject <- rep(paste0("S", 1:n_subjects), each = n_per_subject)
x <- rep(seq(0, 10, length.out = n_per_subject), times = n_subjects)

# subject-level varying effects
intercepts <- rnorm(n_subjects, 0, 1)
slopes <- rnorm(n_subjects, 0, 0.3)
wiggles <- rnorm(n_subjects, 0, 0.2)
subject_df <- tibble(
    subject = paste0("S", 1:n_subjects),
    intercept = intercepts,
    slope = slopes,
    wiggle = wiggles
    )

df <- tibble(x = x, subject = subject) %>%
    left_join(subject_df, by = "subject") %>%
    mutate(
        subject = as.factor(subject),
        y = sin(x) + intercept + slope * x + wiggle * sin(2 * x) + rnorm(n(), sd = 0.3)
        )

######################
# fitting the models #
######################

# fixed-effect GAM
gam_fixed <- gam(
    y ~ s(x),
    data = df
    )

# random intercept + slope
gam_int_slope <- gam(
    y ~ s(x) +
        s(subject, bs = "re") +
        s(x, subject, bs = "re"),
    data = df
    )

# random smooth per subject + global smooth
gam_rand_smooth <- gam(
    y ~ s(x) +
        s(subject, bs = "re") +
        s(x, subject, bs = "fs", m = 1),
    data = df
    )

# defining plotting functions
plot_basis_functions <- function (df, model) {

    # retrieving design matrix and coefficients
    Xp <- predict(model, type = "lpmatrix")
    coefs <- coef(model)

    # basis dataframe
    basis_df <- as.data.frame(Xp)
    colnames(basis_df) <- paste0("b", 1:ncol(basis_df) )
    basis_df$x <- df$x
    basis_long <- pivot_longer(basis_df, starts_with("b"), names_to = "basis", values_to = "value")

    # matching basis to weights
    weights <- data.frame(
        basis = paste0("b", seq_along(coefs) ),
        weight = coefs
        ) %>%
        mutate(label = paste0(basis, " × ", sprintf("%.2f", weight) ) )

    # adding labels to the end of each basis line
    label_df <- basis_long %>%
        group_by(basis) %>%
        slice_tail(n = 1) %>% # get last x point per basis
        left_join(weights, by = "basis") %>%
        mutate(label = paste0(basis, " × ", sprintf("%.2f", weight) ) )

    # fitted curve at the group level
    df$fit <- as.vector(Xp %*% coefs)

    # plotting everything
    ggplot() +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_line(
            # data = basis_long, aes(x = x, y = value, color = basis),
            data = basis_long, aes(x = x, y = value, group = basis),
            colour = "grey50",
            alpha = 0.5,
            linewidth = 0.5
            ) +
        geom_line(
            data = df,
            aes(x = x, y = fit),
            color = "black",
            linewidth = 1.0
            ) +
        # geom_text(
        #     data = label_df,
        #     aes(x = x, y = value, label = label, color = basis),
        #     hjust = 0, nudge_x = 0.1, size = 2
        #     ) +
        # ggrepel::geom_text_repel(
        #     data = label_df,
        #     aes(x = x, y = value, label = label, color = basis),
        #     hjust = 0, nudge_x = 1,
        #     size = 2, direction = "y"
        #     ) +
        labs(x = "Predictor", y = "Response") +
        ylim(-2, +2) +
        # scale_colour_manual(values = met.brewer(name = "Johnson", n = 10) ) +
        theme(legend.position = "none")

plot_model <- function (df, model) {

    # predictions fo the full model: includes all terms (fixed + random)
    df$fit_subject <- predict(model, newdata = df)

    # identify which smooth terms to exclude (those involving 'subject')
    smooth_labels <- sapply(model$smooth, function(s) s$label)
    subject_terms <- smooth_labels[grepl("subject", smooth_labels)]

    # group-level prediction + standard error (excludes subject-specific terms)
    pred_global <- predict(model, newdata = df, exclude = subject_terms, se.fit = TRUE)
    df$fit_global <- pred_global$fit
    df$fit_global_upper <- pred_global$fit + pred_global$se.fit
    df$fit_global_lower <- pred_global$fit - pred_global$se.fit

    # plotting the predictions
    df %>%
        ggplot(aes(x = x, y = y, colour = subject) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_point(
            size = 0.5,
            show.legend = FALSE
            ) +
        geom_line(
            aes(y = fit_subject, group = subject),
            alpha = 0.5,
            linewidth = 0.5,
            show.legend = FALSE
            ) +
        geom_ribbon(
            aes(x = x, y = y, ymin = fit_global_lower, ymax = fit_global_upper),
            fill = "black",
            alpha = 0.2,
            inherit.aes = FALSE
            ) +
        geom_line(
            aes(y = fit_global, group = 1),
            color = "black",
            linewidth = 1
            ) +
        ylim(-5, 5) +
        scale_colour_manual(values = met.brewer(name = "Johnson", n = 5) ) +
        labs(x = "Predictor", y = "Response")

# plotting, note that the basis plot uses the fixed-effect model for simplicity
p1 <- plot_basis_functions(df, gam_fixed)
p2 <- plot_model(df, gam_fixed)
p3 <- plot_model(df, gam_int_slope)
p4 <- plot_model(df, gam_rand_smooth)

# combining plots with labels
(p1 | p2) / (p3 | p4) + plot_annotation(tag_levels = "A")
```

See @fig-intro-gam... Introduction to multilevel GAMs [@pedersen_hierarchical_2019]... Now describe the Bayesian GAMM [@miller2025]... Proper inclusion of varying/random effects in the model specification protects against overly wiggly curves [@baayen2020]... Generalising to scale and shape or "distributional GAMs" [@rigby2005; @umlauf2018] and applied to neuroimaging data [@dinga2021]...

@@

Given the previously reported limitations of conventional methods to precisely identify the onset and offset of M/EEG effects (e.g., ERPs, decoding performance), we developed a model-based approach for estimating the onset and offset of such effects. To achieve this, we leveraged Bayesian generalised additive multilevel models (BGAMMs) fitted in `R` via the `brms` package and compared the performance of this approach to conventional methods on both simulated and actual M/EEG data.

@@

Following the approach of @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see code in the online supplementary materials and details in @yeung2004]. As in @rousselet_using_2025, the signal was generated from a truncated Gaussian distribution with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated data for a group of 20 participants (with variable true onset) with 50 trials per participant and condition (@fig-eeg).

@@

@@

# plotting the data
raw_df %>%
    summarise(
        eeg_mean = mean(eeg),
        eeg_se = sd(eeg) / sqrt(n() ),
        .by = c(participant, condition, time),
        ) %>%
    ggplot(aes(x = time, y = eeg_mean, colour = condition, fill = condition) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_ribbon(
        aes(ymin = eeg_mean - eeg_se, ymax = eeg_mean + eeg_se, colour = NULL),
        alpha = 0.3, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~participant) +
    scale_colour_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "Simulated EEG signal (a.u.)")
```

## Model description and model fitting

We then fitted a Bayesian GAM (BGAM) using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms` (i.e., weakly informative priors). We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$ [@gabry2019; @gelman2020]. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects. @fig-plot-post-slope shows the predictions of this model together with the raw data.

@@

# fitting the BGAM
gam <- brm(
    # cubic regression splines with k-1 basis functions
    # eeg ~ condition + s(time, bs = "cr", k = 20, by = condition),
    # thin-plate regression splines with k-1 basis functions
    eeg ~ condition + s(time, bs = "tp", k = 20, by = condition),
    data = ppt_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/gam.rds"
    )
```

However, the previous model only included constant (fixed) effects, thus not properly accounting for between-participant variability. We next fit a multilevel version of the BGAM (BGAMM). Although it is possible to fit a BGAMM at the single-trial level, we present a resource-efficient version of the model that is fitted directly on the by-participant summary statistics (mean and SD), similar to what is done in meta-analysis.

```{r meta-gam, results = "hide"}
# averaging across participants
summary_df <- raw_df %>%
    summarise(
        eeg_mean = mean(eeg),
        eeg_sd = sd(eeg),
        .by = c(participant, condition, time)
        )

# fitting the BGAMM
meta_gam <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 20, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/meta_gam.rds"
    )
```

@@

We depict the posterior predictions together with the posterior estimate of the slope for `condition` at each timestep (@fig-plot-post-slope). This figure suggests that the BGAMM provides an adequate description of the simulated data.

```{r fig-plot-post-slope, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior estimate of the EEG activity in each condition (left) and posterior estimate of the difference in EEG activity (right) according to the BGAMM."}
# defining a function to plot posterior predictions
plot_post_preds <- function (model, data = NULL) {

@@

    post_preds %>%
        ggplot(
            aes(
                x = time,
                y = eeg_mean,
                colour = condition, fill = condition
                )
            ) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_vline(xintercept = true_offset, linetype = 2) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_point(
            data = data %>% summarise(eeg_mean = mean(eeg_mean), .by = c(time, condition) ),
            aes(fill = NULL),
            pch = 21, show.legend = FALSE
            ) +
        geom_line(aes(y = estimate__), show.legend = FALSE) +
        geom_ribbon(
            aes(ymin = lower__, ymax = upper__, colour = NULL),
            alpha = 0.25, show.legend = FALSE
            ) +
        scale_colour_met_d(name = "Johnson") +
        scale_fill_met_d(name = "Johnson") +
        labs(x = "Time (s)", y = "EEG signal (a.u.)")

# defining a function to plot posterior slope
plot_post_slope <- function (model, data = NULL) {
